Reasoning Graphically about Quantum Computation

### 2008-03-04 papers/reasoning-graphically/paper.tex
pollishing..
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1398 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

  \keywords{graph rewriting, quantum computing, categorical
    logic, interactive theorem proving, graphical calculi}
\end{abstract}

@@

A \emph{directed graph}\footnote{ Equivalently: a directed
  graph is a functor $G$ from $\bullet \pile{\rTo\\\rTo} \bullet$ to
  \catSet; a graph morphism is then a natural transformation $f: G
  \Rightarrow H$.  } consists of a 4-tuple $(V,E,s,t)$ where $V$ and
$E$ are sets, respectively of \emph{vertices}\footnote{We will
  use the words ``vertex'' and ``node'' interchangeably.} and
\emph{edges}, and $s$ and $t$ are maps
\begin{diagram}
  E & \pile{\rTo^{\qquad s\qquad}\\\rTo_t} & V
\end{diagram}
which we call \emph{source} and \emph{target}.  The definitions
$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ express the
\emph{incoming} and \emph{outgoing} edges at a vertex $v$.  The
\emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
\sizeof{\text{out}(v)}$.

To distinguish between elements of different graphs, we will use the
subscript notation $G = (V_G,E_G,s_G,t_G)$. Given graphs $G$ and $H$,
a \emph{graph morphism} $f : G\to H$ consists of functions $f_E : E_G
\to E_H$ and $f_V:V_G\to V_H$ such that:

@@

Let $f: G \to H$ be graph morphism and  let $V' \subseteq V_G$. We say
that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
that $f_E(e') = e$.

\begin{definition}
\label{open-embedding-def}
We call $f$ an \emph{open embedding} for $V'$ if:
\begin{enumerate}
\item $f_E$ is injective;
\item $f_V$ restricted to $V'$ is injective; and,
\item $f$ is strict for $V'$.
\end{enumerate}
\end{definition}

@@

Every arrow $f:A\to B$ in a compact closed category \catC
has a \emph{name} and \emph{coname}:
\[
\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
\]
are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
$\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
\catC(A\otimes B^*,I)$ making \catC monoidally closed\footnote{In
  general compact closed categories  are models of multiplicative
  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
\[
f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
\id{B^*}) \circ (d_A \otimes \id{B^*})
\]
By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
f$.  Thus $(\cdot)^*$ lifts to an involutive functor
$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
opposite.

@@

\section{Graphs as Quantum Computations}
\label{sec:quotients-rewriting}

@@

\begin{theorem}
  Let $G$ be a connected graph generated from $\delta_Z$, $\epsilon_Z$,
  $\alpha_Z$ and their adjoints;  then $G$ is totally determined by
  the number of inputs, the number of outputs, and the sum modulo
  $2\pi$ of the $\alpha$s which occur in it.
\end{theorem}

Hence any connected subgraph
involving nodes of only one colour, wherever it occurs in a graph, may
be collapsed to a single vertex labelled by a single value $\alpha$,
giving a ``spider'' as depicted below:
\[
\inlinegraphic{3em}{gen-spider}
\]
Conversely, a spider may be arbitrarily divided into subspiders,
provided the total in- and out-degree is preserved, along with and the
sum of the $\alpha $s.  Further,
one can derive from the  spider theorem $n$-fold versions of many of
the other rewrites in the system.

@@

\noindent However, composition with semi-circles ($d$ and $e$ from
Definition~\ref{compactcat-def}) on the co-domain or domain allows an
equation involving one of the above to easily lead to a derivation
corresponding either of the other two. For instance, proving any of
the following allows a trivial derivation of the others:

@@

\noindent This motivates a characterisation that has only one
representative for such a set of equations. More presicely, we
formalise a representation that abstracts over the boundary nodes
membership in the domain or co-domain. This gives rise to a
\emph{variable-node graphs}, in which boundary nodes have been
generalised to \emph{variable nodes}. The intuition of variable nodes
is that they can replaced by concrete nodes in some graph. We now
define the semantics for variable-node graphs and then go on and
introduce a way to join them.

@@

To formalise the semantics of variable-node graphs, we define
matching, which captures the intuitive idea of graph with variable
nodes occuring within another graph.

@@

A graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs, denoted by
$\vinterp{G}$. The interpretation is simply the ways a graph with
variable nodes can be matched to a concrete graph. We refer to members
of this set as instances. Matching is also well-defined between
variable-node graphs, in which case we denote that $G$ macthes $H$ by
$G \leq_v H$.

\begin{theorem}
\label{match-interp-thm}
  $G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. Proof... 
\end{theorem}

\begin{theorem}
\label{reflexive-match-thm}
  \emph{Matching is reflexive}: $G \leq_v G$ is a trivial consequence from open-embedding. 
\end{theorem}

@@

\begin{definition}
  A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of
  strict partial homomorpihc mappings, $f_G : G \Longrightarrow H$ and
  $f_H : H \Longrightarrow G$ such that $f_G$ and $f_H$ do not map
  concrete nodes and for the nodes that are mapped \emph{agree}
  following the definition given below.
\end{definition}

@@

We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
overload the notation $\pi_p(G,H)$ to let it denote the graph that
results from the plugging. This is defined to be the union of the
nodes and edges from $G$ and $H$ where the nodes and edges mapped by
the pair of plugging function are identified. Identification of nodes
instantiates variable nodes: when a variable node is mapped to a
concrete one, identification results in the concrete node.

While this is simply the intuitive way to combine graphs with variable
nodes, the definition takes some care. In particular, it is important
that plugging preserves the types of concrete nodes and results in
subtypes in variable nodes. Figure~\ref{plugging-examples-fig}
presents some examples to help the reader get the intuition as well as
understand some of the more subtlte cases.

\begin{theorem}
  \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
  = \pi_q(H,G)$. Proof follows simply: given $p = (f_G,f_H)$, $q =
  (f_H,f_G)$. As a corrolary, we get $\pi(G,H) \Leftrightarrow
  \pi(H,G)$.
\end{theorem}

@@

A crucial feature of rules like the Spider Law is that an arbirary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing, on graph
representations. Given a graph representation, this introduces a new
notation by outlining (called !-boxing) a set of chosen
nodes. Intuitively, the resulting !-boxed graph can be thought of as
representing a set of graphs with an arbirary number of copies of the
!-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies, which erases all edges to and from the !-box. See
Figure~\ref{spider-bang-rule} .

@@

The representation of Compact Closed Categories as Graphs, discussed
in~\S\ref{graph-repr-comp}, is too restrictive for reasoning about
quantum computation. In particular, pictures such the Spider Law, in
Figure~\ref{spider-law-fig} are frequently needed, but not
expressible. Such pictures are used to denote an infinite family of
concrete graphs that can be rewritten. In this section we combine the
{\em !-boxes} with {\em variable nodes} graph extensions to form a
representation that allows us to finitely express these infinite
families of concrete graphs. We call this extension \ref{graph
  patterns}. The Spider Law can now be represented by the equation in
Figure~\ref{graph-pattern-spider-law-fig}. We also extend the notion
of matching for graph pattens as well as present a corresponding
notion of plugging. This provides the foundations for the rewriting
machinery in~\S\ref{rewriting} which can then be used to reasoning
about quantum computation.

>>>>

  \keywords{graph rewriting, quantum computing, categorical logic,
    interactive theorem proving, graphical calculi; concerning:
    symbolic computation, automated reasoning, formal mathematics,
    logic}
\end{abstract}

@@

A \emph{directed graph}\footnote{ Equivalently: a directed
  graph is a functor $G$ from $\bullet \pile{\rTo\\\rTo} \bullet$ to
  \catSet; a graph morphism is then a natural transformation $f: G
  \Rightarrow H$.  } consists of a 4-tuple $(V,E,s,t)$ where $V$ and
$E$ are sets, respectively of \emph{vertices}\footnote{We will
  use the words ``vertex'' and ``node'' interchangeably.} and
\emph{edges}, and $s$ and $t$ are maps
\begin{diagram}
  E & \pile{\rTo^{\qquad s\qquad}\\\rTo_t} & V
\end{diagram}
which we call \emph{source} and \emph{target}.  The definitions
$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ express the
\emph{incoming} and \emph{outgoing} edges at a vertex $v$.  The
\emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
\sizeof{\text{out}(v)}$. To distinguish between elements of different
graphs, we will use the subscript notation $G = (V_G,E_G,s_G,t_G)$.

Given graphs $G$ and $H$, a \emph{graph morphism} $f : G\to H$
consists of functions $f_E : E_G \to E_H$ and $f_V:V_G\to V_H$ such
that:

@@

\noindent These ensure that the structure of the graph is preserved by
the morphism: an edge connected to a node gets mapped to a new edge
that must connected in the same way to the mapped node.

Let $f: G \to H$ be graph morphism and let $V' \subseteq V_G$. We say
that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
that $f_E(e') = e$. Strictness ensures that there are no additional
edges in $H$.

\begin{definition}
\label{open-embedding-def}
We call a graph morphism $f$ an \emph{open embedding} for $V'$ if:
\begin{enumerate}
\item $f_E$ is injective;
\item $f_V$ restricted to $V'$ is injective; and,
\item $f$ is strict for $V'$.
\end{enumerate}
\end{definition}

@@

Every arrow $f:A\to B$ in a compact closed category \catC
has a \emph{name} and \emph{coname}:
\[
\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
\]
which are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
$\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
\catC(A\otimes B^*,I)$ making \catC monoidally closed\footnote{In
  general compact closed categories  are models of multiplicative
  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
\[
f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
\id{B^*}) \circ (d_A \otimes \id{B^*})
\]
By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
f$.  Thus $(\cdot)^*$ lifts to an involutive functor
$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
opposite.

@@

\section{Quantum Computations as Graphs}
\label{sec:quotients-rewriting}

@@

\begin{theorem}
  The \emph{Spider Theorem}: let $G$ be a connected graph generated
  from $\delta_Z$, $\epsilon_Z$, $\alpha_Z$ and their adjoints; then
  $G$ is totally determined by the number of inputs, the number of
  outputs, and the sum modulo $2\pi$ of the $\alpha$s which occur in
  it.
\end{theorem}

Hence any connected subgraph
involving nodes of only one colour, wherever it occurs in a graph, may
be collapsed to a single vertex labelled by a single value $\alpha$,
giving a ``spider'' as depicted below:
\[
\inlinegraphic{3em}{gen-spider}
\]
Conversely, a spider may be arbitrarily divided into subspiders,
provided the total in- and out-degree is preserved, along with and the
sum of the $\alpha $s.  Further, one can derive from the spider
theorem $n$-fold versions of many of the other rewrites in the system.
Informally, this can be expressed as the graph equation in
Figure~\ref{informal-spider-law-fig}.

@@

\noindent However, composition with semi-circles ($d$ and $e$ from
Definition~\ref{compactcat-def}), on the co-domain or domain, allows
an equation involving one of the above to easily lead to a derivation
corresponding to either of the other two. For instance, proving any of
the following allows a trivial derivation of the others:

@@

\noindent This motivates a characterisation that has only one
representative for such a set of equations. More presicely, we
formalise a representation that abstracts over the boundary nodes
membership in the domain or co-domain. This gives rise to a
\emph{variable-node graphs}, in which boundary nodes have been
generalised to \emph{variable nodes}. The intuition of variable nodes
is that they can replaced by concrete nodes in some graph. This is a
process analgous to composition. We now define the semantics for
variable-node graphs and then go on and introduce a way to join them.

@@

To formalise the semantics of variable-node graphs, we define
matching, which captures the intuitive idea of a graph with variable
nodes occuring within another graph.

@@

A graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs, denoted by
$\vinterp{G}$. The interpretation is simply the set of concrete graphs
which the variable node graph matches. We refer to members of this set
as instances. Matching is also well-defined between two variable-node
graphs, in which case we denote that $G$ macthes $H$ by $G \leq_v H$.

\begin{theorem}
\label{match-interp-thm}
$G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. The
proof from left to right is a consequence of the trivial lemma that a
compositon of open embeddings is an open embedding. This makes the
embedding of $G$ into $H$ composed with $H$ into $\vinterp{H}$ an open
embedding of $G$ into $\vinterp{H}$. From right to left, we compose
the open embedding $\vinterp{G}$, restricted to it's subset
$\vinterp{H}$, with the inverse of $\vinterp{H}$ to get $G \leq_v H$.
\end{theorem}

\begin{theorem}
\label{reflexive-match-thm}
\emph{Matching is reflexive}: $G \leq_v G$ is a trivial instance of
the same theorem for open-embeddings, where the embedding is the
identity map.
\end{theorem}

@@

\begin{definition}
  A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of strict
  partial graph morphisms, $f_G : G \rightarrow H$ and $f_H : H
  \rightarrow G$ such that $f_G$ and $f_H$ \emph{agree}, following
  the definition given below, on node mappings and do not map concrete
  nodes.
\end{definition}

@@

We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
overload the notation $\pi_p(G,H)$ to denote the graph that results
from the plugging. This is defined to be the union of the nodes and
edges from $G$ and $H$ where the nodes and edges mapped by the pair of
plugging function are identified. Identification of nodes instantiates
variable nodes: when a variable node is mapped to a concrete one,
identification results in the concrete node.

While this is simply the intuitive way to combine graphs with variable
nodes, the definition takes some care. In particular, it is important
that plugging preserves incident edges of concrete nodes.
Figure~\ref{plugging-examples-fig} presents some examples to help the
reader get the intuition as well as understand some of the more
subtlte cases.

\begin{theorem}
  \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
  = \pi_q(H,G)$. Proof, given $p = (f_G,f_H)$ then $q = (f_H,f_G)$. As
  a corollary, we get $\pi(G,H) \Leftrightarrow \pi(H,G)$.
\end{theorem}

@@

A crucial feature of rules like the Spider Theorem is that an arbirary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing, on graph
representations. Given a graph representation, this introduces a new
notation by outlining (called !-boxing) a set of chosen
nodes. Intuitively, the resulting !-boxed graph can be thought of as
representing a set of graphs with an arbirary number of copies of the
!-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies, which erases all edges to and from the !-box. See
Figure~\ref{spider-bang-rule} .

@@

The representation of Compact Closed Categories as Graphs, discussed
in~\S\ref{graph-repr-comp}, is too restrictive for reasoning about
quantum computation. In particular, pictures such the Spider Theorem, in
Figure~\ref{spider-law-fig} are frequently needed, but not
expressible. Such pictures are used to denote an infinite family of
concrete graphs that can be rewritten. In this section we combine the
{\em !-boxes} with {\em variable nodes} graph extensions to form a
representation that allows us to finitely express these infinite
families of concrete graphs. We call this extension \ref{graph
  patterns}. The Spider Law can now be represented by the equation in
Figure~\ref{graph-pattern-spider-law-fig}. We also extend the notion
of matching for graph pattens as well as present a corresponding
notion of plugging. This provides the foundations for the rewriting
machinery in~\S\ref{rewriting} which can then be used to reasoning
about quantum computation.
